---
title: "Manuscript draft: DTAtools (Nature Methods format)"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Manuscript draft: DTAtools (Nature Methods format)}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 80
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

<!--
This vignette is a manuscript draft in the structure of a Nature Methods
Brief Communication / Article. It is a working document, not a submission:

  * The RESULTS benchmark subsections and the DISCUSSION are placeholders.
    They are marked with "[TO BE FILLED IN ...]" and their code chunks are
    `eval = FALSE`. Fill them in once the benchmark campaign has run.
  * Everything else (Abstract, Introduction, Methods, availability
    statements) is prose that should survive into the submission largely
    unchanged, modulo journal-specific word limits.

Nature Methods Brief Communication limits, for reference while editing:
~3 pages, title <= 10 words, abstract <= 70 words (3 sentences),
main text ~1000-1500 words (no headings in print; kept here for drafting),
2 display items (figures/tables), Online Methods unlimited.
-->

# Title

**DTAtools: declarative validation of clinical data transfers in R**

Thomas Schwarzl^1,\*^ and Daniel Schreyer^1^

^1^ Boehringer Ingelheim, Global Computational Biology and Digital Sciences.
^\*^ Correspondence: thomas.schwarzl\@boehringer-ingelheim.com

# Abstract

<!-- Nature Methods Brief Communication: <= 70 words, 3 sentences, no
     references. Current draft is intentionally slightly over; trim at
     submission. -->

Data transfers between research organisations, vendors, and sponsors fail
routinely because delivered files silently deviate from the agreed
specification. DTAtools encodes Data Transfer Agreements as human-readable
YAML and validates tabular deliveries against them in R — column types,
nullability, controlled vocabularies, regular expressions, and cross-column
logic — reporting every deviation on three independent axes. A streaming
validation mode checks files larger than memory with a bounded working set.

# Introduction

Regulated research runs on files that change hands. A contract research
organisation transmits laboratory measurements to a sponsor; a genomics vendor
delivers sample manifests to a pharma company; a biomarker lab returns derived
endpoints for a clinical study. Each transfer is governed by a **Data Transfer
Agreement (DTA)** and its technical annex, the **Data Transfer Specification
(DTS)**: the column names, types, controlled vocabularies, and inter-column
rules the delivered tables must satisfy.

In practice these specifications live in Word documents, and conformance is
checked by hand — or discovered downstream, when an analysis pipeline breaks on
`"unknown"` in a numeric column. Manual review is error-prone and unrepeatable;
every failed delivery costs a resubmission cycle measured in days to weeks.
General-purpose validation frameworks exist for R — `pointblank`, `validate`,
and `arsenal` among them — but none speaks the language of a DTA: none pairs a
document-shaped, signable specification with programmatic enforcement of
exactly that specification, and none addresses deliveries too large to load
into memory.

DTAtools closes this gap. The agreement itself — metadata, contacts,
transmission schedule, and one or more dataset specifications — is a single
YAML document that is at once human-readable, diff-able under version control,
and directly executable as a validation program. One function call reads a
delivered file *in the types the specification declares* and reports every
deviation with row, column, offending value, and violated constraint. The same
YAML renders to a formatted Word or PDF agreement for signature, so the
document the parties sign and the program the machines run can never drift
apart. Validation scales from interactive checks of small tables to a
streaming scan of files larger than memory, and a bundled Shiny application
gives non-programmers the identical validation engine in a browser.

# Results

## A single YAML document is both the agreement and the validation program

A DTA in DTAtools is one YAML file with a `metadata:` block (title, version,
receiver and supplier contacts, transmission schedule) and a `datasets:` list.
Each dataset declares its expected files, per-column specifications (SAS-style
`type`, `nullable`, optional `values`, `pattern`, `length`, `format`), and
cross-column rules. Reading the file yields an S7 object graph mirroring that
structure; validating a delivery is three calls:

```{r quickstart}
library(DTAtools)

dta <- read_dta_from_yaml(
  system.file("extdata", "clinical_dta.yaml", package = "DTAtools")
)
dta <- load_file(
  dta,
  dataset = "clinical_data",
  file = system.file("extdata", "clinical_data.csv", package = "DTAtools")
)
dta <- check(dta, quiet = TRUE)

results(dta)
```

Validation reports on three independent axes, and a failure on any one of them
fails the table: **schema** errors (a value breaks a column constraint),
**rule** errors (a row breaks inter-column logic), and **import** errors (a
value cannot be represented in the type its column declares — the text
`"unknown"` in a numeric column, for example). Import errors are the class of
defect that silent type coercion hides; DTAtools surfaces them as first-class
failures, retaining the original text for inspection:

```{r import-errors}
specs <- DTAColumnSpecCollection(columns = list(
  SUBJECT_ID = DTAColumnSpec(id = "SUBJECT_ID", type = "SAS Char", nullable = FALSE),
  AGE        = DTAColumnSpec(id = "AGE", type = "SAS Num", nullable = TRUE)
))

ds <- DTADataSetTabular(
  name   = "import_demo",
  specs  = specs,
  tables = list(demo = data.frame(
    SUBJECT_ID = c("SUBJ0001", "SUBJ0002"),
    AGE        = c("34", "unknown")
  ))
)
ds <- check(ds, quiet = TRUE)

validation_status(ds)[, c("ok", "n_columnspec_errors", "n_rule_errors", "n_import_errors")]
```

Every reported error carries an `id`; `inspect()` resolves it to the full
story — the row in context, the JSON Schema constraint that failed, or the
original text of a value that would not import. The same objects export the
signable agreement (`write_dta()`, Word/PDF/Markdown), a self-contained HTML
validation report (`write_validation_report()`), and validated data with MD5
checksums (`write_table_to_file()`).

## Rules capture the cross-column logic of real agreements

Beyond per-column constraints, four declarative rule types cover the
inter-column logic that real DTAs contain: `col_condition` (row-wise if/then),
`col_range`, `col_unique` (single or composite key), and `group_condition`,
which groups rows by one or more key columns and enforces constraints across
each whole group — "if any row for a subject records consent, every row for
that subject must carry a consent date". Rules are written in the same YAML
document, evaluated by the same `check()` call, and reported on the same
`messages()`/`inspect()` surface as column errors.

## Validation of files larger than memory

`load_file(stream = "always")` holds a delivery as a lazy Arrow Dataset
instead of an in-memory table; `check()` then scans it in batches, and
`validate_file_stream()` performs the same scan without constructing the
object model at all. Nothing in the scan grows with the number of rows: memory
is bounded by the batch size for column checks, by the number of distinct keys
for uniqueness rules, and by the number of distinct groups for grouped rules.
Both paths find the same errors and reach the same verdict.

<!-- ============================================================ -->
<!-- BENCHMARK PLACEHOLDER — the subsections below are scaffolds. -->
<!-- Run the benchmark campaign, then replace the bracketed        -->
<!-- placeholders with measured numbers and set the chunks to      -->
<!-- eval = TRUE (or paste in pre-computed result tables).         -->
<!-- ============================================================ -->

## Benchmark: throughput and memory across delivery sizes

**[TO BE FILLED IN — benchmark results.]** *This subsection will report
wall-clock time, CPU time, R heap peak, and Arrow pool peak for the in-memory
and streaming validation paths across delivery sizes spanning [N] orders of
magnitude ([sizes to be specified], uncompressed and gzip-compressed), on
[hardware/OS to be specified]. Metrics are collected with the package's
built-in instrumentation (Methods); the exact code is below, so filling in
this section is a matter of running it on the benchmark machine and pasting
the resulting table.*

```{r benchmark-scaffold, eval = FALSE}
# Scaffold for the benchmark campaign. Point `files` at the generated
# delivery files (Methods) and run on the benchmark machine.
files <- c(
  # "delivery_1e5_rows.csv", "delivery_1e6_rows.csv", ...
)

bench_one <- function(file, stream) {
  dta <- read_dta_from_yaml("benchmark_dta.yaml")
  dta <- load_file(dta, dataset = 1, file = file, stream = stream)
  dta <- check(dta, quiet = TRUE, benchmark = TRUE)
  cbind(file = basename(file), stream = stream, validation_benchmark(dta))
}

benchmarks <- do.call(rbind, c(
  lapply(files, bench_one, stream = "never"), # in-memory path
  lapply(files, bench_one, stream = "always") # streaming path
))
benchmarks
```

**Placeholder for Figure 2** *(throughput and peak memory vs. delivery size,
in-memory vs. streaming; one panel per metric)*: **[TO BE FILLED IN once the
benchmark table above exists.]**

## Benchmark: comparison with existing validation frameworks

**[TO BE FILLED IN — comparative benchmark results.]** *This subsection will
compare DTAtools with [`pointblank`, `validate`, and other frameworks to be
finalised] on an equivalent constraint set: features first (which DTA
constraint classes each framework can express at all — declared column types
at read time, import-error detection, grouped cross-row rules, larger-than-
memory input), then runtime and memory on the subset of constraints every
framework supports. The feature matrix is drafted in the Methods; the numbers
await the campaign.*

# Discussion

**[TO BE FILLED IN — conclusion depends on the benchmark outcome.]** *Drafting
notes for the eventual text: interpret the benchmark results (does the
streaming path hold its bounded-memory promise at every tested size? where is
the throughput crossover between the in-memory and streaming paths? how does
constraint evaluation cost compare with parsing cost?); position DTAtools
against the compared frameworks based on the measured feature and performance
matrix, not on claims; state the limitations that the results reveal; close
with the practice implication — specification-first validation at the point of
transfer, before submission rather than after failure.*

# Methods

## Software architecture

DTAtools is an R package (R ≥ 4.1.0) built on the S7 object system. A
top-level `DTA` object holds a `DTAMetaData` (title, version, contacts,
transmission schedule) and a named list of datasets. Two dataset classes
implement the two kinds of deliverable found in agreements:
`DTADataSetTabular` carries file handlers, a `DTAColumnSpecCollection`
(per-column `DTAColumnSpec` objects plus a list of `DTARule` objects), and the
delivered tables as Arrow Tables; `DTADataSetFile` checks only that declared
files arrived, are non-empty, and are readable — for PDFs, archives, and raw
instrument output that are never parsed. File handlers (`DTAFileCSV`,
`DTAFileTSV`, `DTAFileDelim`, `DTAFileAny`) encapsulate delivery-name
matching — including filename regular expressions, expected file counts, and
transparent acceptance of gzip-compressed uploads — and reading.

## Typed import and the three validation axes

A column's declared SAS-style type (`SAS Char`, `SAS Num`, `SAS Int`,
`SAS Date`, `SAS Time`, `SAS DateTime`) is applied when the data is *read*,
not merely when it is validated: `SAS Char` pins a column to text (so `"007"`
stays `"007"`), numeric types store numbers, and date/time types keep the text
as written and validate it by pattern. A source value that cannot be
represented in its declared type becomes `NA`, and the event is recorded as an
**import error** carrying the original text and the reason — a validation
failure in its own right, independent of the schema and rule axes. Column
constraints are compiled to JSON Schema and evaluated row-wise; rule
evaluation runs after column validation on the typed table. `check()` skips
tables whose data and specifications are unchanged since the last run, and
`force = TRUE` overrides.

## Streaming validation

For deliveries larger than memory, `check()` on a lazily-opened dataset and
`validate_file_stream()` scan the file in batches of
`getOption("DTAtools.stream_batch_rows", 131072L)` rows (Arrow record
batches), evaluating column constraints per batch and maintaining bounded
state across batches: a distinct-key map for uniqueness rules, per-group flag
aggregates for grouped rules, and a `max_errors` cap on retained per-cell
detail. None of these grow with row count. `fail_fast = TRUE` stops at the
first problem and marks the result as an explicitly partial scan, reporting
`NA` rather than `TRUE` for any axis it could not settle.

## Benchmark instrumentation

All timings and memory figures **[will be]** collected with the package's
built-in instrumentation: `check(benchmark = TRUE)` and
`validate_file_stream(benchmark = TRUE)` attach a one-row metrics frame
retrievable with `validation_benchmark()`, reporting elapsed and CPU seconds
(`proc.time()`), R heap peak (via `gc(reset = TRUE)` at call start), process
RSS at start and end, and Arrow memory-pool figures. Because Arrow's default
memory pool exposes only a session-lifetime high-water mark,
`arrow_pool_peak_mb` reports that absolute peak, `arrow_call_mb` the
difference observed across the call, and `arrow_call_exact` whether this call
actually set a new high-water mark — when `FALSE`, `arrow_call_mb` is a lower
bound, not a measurement. On Windows, `proc.time()` does not measure
child-process CPU, so cross-platform CPU comparisons use the elapsed column.

**[TO BE FILLED IN — benchmark design.]** *Hardware, OS, R and package
versions, Arrow version and thread settings, delivery-file generator (row
counts, column mix, error injection rates), number of repetitions, and
warm/cold cache protocol — to be recorded from the actual campaign. The
comparative feature matrix (which constraint classes each compared framework
expresses) is to be finalised together with the framework list.*

## Data availability

The package bundles a complete example agreement
(`inst/extdata/clinical_dta.yaml`) and matching example deliveries, used
throughout this manuscript. **[Benchmark input files are generated
synthetically by the script referenced above; the generator and its seed will
be published with the benchmark results.]**

## Code availability

DTAtools is open source under LGPL (≥ 3) at
<https://github.com/Boehringer-Ingelheim/DTAtoolsR>. All analyses in this
manuscript are contained in the package vignettes and run with
`R CMD check`.

## Author contributions

T.S. conceived the project and wrote the package. D.S. contributed to the
package. **[To be completed and confirmed by all authors.]**

## Competing interests

Both authors are or were employees of Boehringer Ingelheim.

# References

<!-- Placeholder reference list — replace with the journal's citation
     format and a curated bibliography at submission. -->

1. R Core Team. *R: A Language and Environment for Statistical Computing*
   (R Foundation for Statistical Computing).
2. Richardson, N. et al. *arrow: Integration to Apache Arrow*
   (CRAN package).
3. Vaughan, D., Wickham, H. & Henry, L. *S7: An Object Oriented System Meant
   to Become a Successor to S3 and S4* (CRAN package).
4. Iannone, R. *pointblank: Data Validation and Organization of Metadata for
   Local and Remote Tables* (CRAN package).
5. van der Loo, M. & de Jonge, E. Data validation infrastructure for R.
   *J. Stat. Softw.* **97**, 10 (2021).
6. **[Further references to be added with the Results.]**

# Session information

```{r session-info}
sessionInfo()
```
