---
title: "DTAtools: Validating Data Transmission Specifications in R"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{DTAtools: Validating Data Transmission Specifications in R}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 80
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = TRUE
)
```

## Introduction

`DTAtools` automates the validation of tabular data against **Data Transmission
Agreement (DTA)** and **Data Transmission Specification (DTS)** definitions. In
regulated industries — clinical research, biomarker science, genomics — data
must conform precisely to agreed column types, allowed values, and cross-column
logic before it is transmitted between parties. Catching these errors
programmatically, before submission, saves costly re-submissions and ensures
reproducible, auditable data exchange.

`DTAtools` stores specifications in human-readable **YAML** files and provides
an R API to:

1. **Import** column specifications and semantic rules from YAML (or Word documents).
2. **Validate** tabular data against those specifications — types, nullability, allowed values, regex patterns, and cross-column rules.
3. **Inspect** validation results programmatically: which tables failed, which rows, which columns, which rules — and drill into individual errors with `inspect()`.
4. **Export** validated data to files (TSV/CSV, optionally gzip-compressed, with MD5 checksums).
5. **Generate** Word documentation tables directly from specs — for use inside DTA/DTS documents.

`DTAtools` is built on the S7 object system and uses JSON Schema for fast,
row-level validation.

---

## Installation

```{r, eval=FALSE}
# Install from GitHub
remotes::install_github("Boehringer-Ingelheim/DTAtoolsR")
library(DTAtools)
```

```{r}
library(DTAtools)
```

---

## Package Architecture

`DTAtools` organises objects into a layered hierarchy that maps directly to how
a Data Transmission Specification (DTS) is structured:

```
DTA                              ← top-level agreement container
├── DTAMetaData                  ← title, version, receiver/supplier contacts,
│                                     transmission schedule
└── DTADataSet (named list)      ← one or more datasets per DTS
    │
    ├── DTADataSetTabular        ← tabular data: full column-spec + rule validation
    │   ├── files (list)         ← expected file descriptors
    │   │   └── DTAFileCSV / DTAFileTSV / DTAFileDelim
    │   │       └── read_file()  → Arrow Table
    │   ├── specs                ← column-level schema + cross-column rules
    │   │   └── DTAColumnSpecCollection
    │   │       ├── DTAColumnSpec × N   ← one per column: type, nullable,
    │   │       │                            values/pattern, length, format
    │   │       └── DTARuleCollection
    │   │           ├── DTARuleColCondition × N  ← if/then cross-column logic
    │   │           ├── DTARuleColRange × N      ← numeric range constraint
    │   │           └── DTARuleColUnique × N     ← uniqueness (single or composite key)
    │   └── tables (named list)  ← actual data as Arrow Tables
    │       └── <table_name>     ← populated via load_file() or constructor
    │
    └── DTADataSetFile           ← non-tabular: file presence/readability check
        ├── file_paths           ← character vector of expected file paths
        └── files (list)         ← DTAFile descriptors (derived from file_paths)
```

**Two ways to build any object:**

- **From YAML:** `read_dta_from_yaml()` → full DTA; `import_specs_from_yaml()` → standalone specs  
- **Programmatically in R:** construct each class directly (e.g. `DTAColumnSpec()`, `DTADataSetTabular()`)

**Validation flow:**

1. `load_file(dta, dataset, file)` — reads a file into the dataset using its file handler
2. `check(dta)` or `check(ds)` — runs JSON Schema validation + rule checks; returns the updated object
3. `results(x)` — one-row-per-table summary (pass/fail, error counts)
4. `messages(x)` — one-row-per-error detail table (column, row, rule, message, `id`)
5. `inspect(x, id = N)` — deep-dive into a specific error message by its `id`

---

## Quickstart

The canonical production workflow with a full DTA YAML: load the specification,
load the data file, validate, inspect.

```{r quickstart}
library(DTAtools)

# 1. Load the full DTA definition from YAML
#    (metadata + dataset specs + file handlers + rules — all in one file)
dta <- read_dta_from_yaml(
  system.file("extdata", "clinical_dta.yaml", package = "DTAtools")
)
print(dta)

# 2. Load the data file into the DTA object
#    load_file() uses the file handler already defined in the YAML
csv_path <- system.file("extdata", "clinical_data.csv", package = "DTAtools")
dta <- load_file(dta, dataset = "clinical_data", file = csv_path)

# 3. Validate — check() runs schema + rule validation for all datasets
dta <- check(dta)

# 4. Summarise results (one row per table)
results(dta)
```

---

## Column Specifications

### The `DTAColumnSpec` Object

A `DTAColumnSpec` represents a single column's metadata and constraints. It maps
directly to a column entry in a YAML specification.

| Field         | Required | Description                                               |
|---------------|----------|-----------------------------------------------------------|
| `id`          | ✔        | Column name (must match the data frame header exactly)    |
| `label`       |          | Human-readable label for documentation                   |
| `type`        |          | SAS-style type: `SAS Char`, `SAS Num`, `SAS Int`         |
| `format`      |          | SAS format string, e.g. `SAS $10.`, `SAS 8.`            |
| `length`      |          | Maximum character length (`maxLength` in JSON Schema)     |
| `nullable`    |          | `TRUE`/`FALSE` — whether the column may be empty/NA      |
| `values`      |          | Closed list of allowed values (`enum` in JSON Schema)    |
| `pattern`     |          | Regular expression every value must match                |
| `description` |          | Free-text description for documentation                  |

> **Note:** `values` and `pattern` are mutually exclusive — a column either has
> a closed value list or a regex pattern, not both.

```{r column_spec_manual}
# A column with a closed value list
visit_col <- DTAColumnSpec(
  id          = "VISIT",
  label       = "Visit Code",
  type        = "SAS Char",
  length      = 12,
  nullable    = FALSE,
  values      = list("V01", "V02", "V03", "EOT"),
  description = "Planned study visit code"
)
print(visit_col)

# A column with a regex pattern
subjid_col <- DTAColumnSpec(
  id          = "SUBJECT_ID",
  label       = "Subject Identifier",
  type        = "SAS Char",
  nullable    = FALSE,
  pattern     = "^SUBJ[0-9]{4}$",
  description = "Unique subject ID in format SUBJ0001"
)
print(subjid_col)

# A numeric column with no further constraints
age_col <- DTAColumnSpec(
  id       = "AGE",
  label    = "Age (years)",
  type     = "SAS Num",
  nullable = FALSE
)
print(age_col)

# A nullable column (empty / NA allowed)
ae_col <- DTAColumnSpec(
  id       = "AE_TERM",
  label    = "Adverse Event Term",
  type     = "SAS Char",
  nullable = TRUE
)
print(ae_col)
```

### The `DTAColumnSpecCollection` Object

A `DTAColumnSpecCollection` groups column specs together, optionally with a set
of schema rules. This is the core specs object passed to `DTADataSetTabular`.

```{r spec_collection}
# Build a collection manually
studyid_col <- DTAColumnSpec(
  id = "STUDYID", type = "SAS Char", nullable = FALSE,
  values = list("1234-5678")
)
gender_col <- DTAColumnSpec(
  id = "GENDER", type = "SAS Char", nullable = FALSE,
  values = list("Male", "Female", "Other")
)

specs <- DTAColumnSpecCollection(
  columns = list(
    STUDYID = studyid_col,
    VISIT   = visit_col,
    AGE     = age_col,
    GENDER  = gender_col,
    AE_TERM = ae_col
  )
)
print(specs)

# Access individual column specs
names(specs)
colspec(specs, "VISIT")
```

### Importing Specifications from YAML

The most common way to work with specs is to load them from a YAML file. The
YAML file must contain a `columns` key with a list of column definitions.

```{r import_yaml}
# The package ships with example YAML specs
yaml_file <- system.file("extdata", "gf_dataset.yaml", package = "DTAtools")
gf_specs <- import_specs_from_yaml(yaml_file)
print(gf_specs)
```

The resulting object contains both columns and rules:

```{r inspect_specs}
# How many columns?
length(gf_specs@columns)

# Look at an individual spec by column ID
colspec(gf_specs, "GFSTAT")

# Are there rules?
gf_specs@rules
```

> **Nested YAML:** If your specs live inside a larger YAML file under a nested
> key (e.g. under `DTA: format:`), pass `tag = c("DTA", "format")` to
> `import_specs_from_yaml()`.

### Exporting Specifications to YAML and JSON

Specifications can be serialised back to YAML or JSON, making it easy to share,
version-control, or programmatically generate specs.

```{r export_specs_yaml}
# Write to YAML
yaml_out <- tempfile(fileext = ".yaml")
write_columns_to_yaml(gf_specs, yaml_out)
cat(paste(readLines(yaml_out, n = 20), collapse = "\n"))
unlink(yaml_out)
```

```{r export_specs_json}
# Write to JSON
json_out <- tempfile(fileext = ".json")
write_columns_to_json(gf_specs, json_out)
cat(paste(readLines(json_out, n = 25), collapse = "\n"))
unlink(json_out)
```

---

## Data Validation

### Creating a `DTADataSetTabular`

A `DTADataSetTabular` combines a `DTAColumnSpecCollection` with one or more
data tables. Tables are stored internally as Arrow Tables for performance.

```{r create_dataset}
# Load the clinical DTA specs from a full DTA YAML (metadata + datasets)
clinical_yaml <- system.file("extdata", "clinical_dta.yaml", package = "DTAtools")
dta_obj       <- read_dta_from_yaml(clinical_yaml)

# The DTA contains one dataset called "clinical_data" — but tables are empty
# until you load data into them
datasets(dta_obj)
ds_spec <- datasets(dta_obj, "clinical_data")
ds_spec   # specs defined, no tables yet
```

The most flexible approach is to combine specs with your own data:

```{r load_data_into_ds}
# Get specs from the full DTA spec
clinical_specs <- specs(ds_spec)

# Read the CSV independently
csv_path   <- system.file("extdata", "clinical_data.csv", package = "DTAtools")
raw_data   <- read.csv(csv_path)

# Create the DTADataSetTabular with specs + data, then validate
ds <- DTADataSetTabular(
  name   = "clinical_data",
  specs  = clinical_specs,
  tables = list(clinical_data = raw_data)
)
ds
```

Alternatively, create a dataset directly in R with inline data:

```{r inline_dataset}
# Build specs
col_studyid <- DTAColumnSpec(id = "STUDYID", type = "SAS Char", nullable = FALSE,
                             values = list("STUDY001"))
col_visit   <- DTAColumnSpec(id = "VISIT",   type = "SAS Char", nullable = FALSE,
                             values = list("SCREENING", "BASELINE", "WEEK4"))
col_age     <- DTAColumnSpec(id = "AGE",     type = "SAS Num",  nullable = FALSE)

inline_specs <- DTAColumnSpecCollection(
  columns = list(STUDYID = col_studyid, VISIT = col_visit, AGE = col_age)
)

# Build data
inline_data <- data.frame(
  STUDYID = c("STUDY001", "STUDY001", "STUDY001"),
  VISIT   = c("SCREENING", "BASELINE", "WEEK4"),
  AGE     = c(32, 32, 32)
)

# Combine into a DTADataSetTabular
inline_ds <- DTADataSetTabular(
  name   = "inline_example",
  specs  = inline_specs,
  tables = list(my_table = inline_data)
)
inline_ds
```

### Validating with `check()`

`check()` runs two validation passes:

1. **JSON Schema validation** — checks every row against column types,
   `nullable`, `values`, `pattern`, and `length`.
2. **Schema rule validation** — evaluates inter-column rules (`col_condition`,
   `col_range`, `col_unique`).

`check()` **always returns the updated object** — assign the result back.

#### Validating a dataset

```{r validate_ds}
ds <- check(ds)
```

### Inspecting Validation Results

After `check()`, use these functions to inspect what happened:

| Function              | Returns                                                   |
|-----------------------|-----------------------------------------------------------|
| `results(x)`          | One row per table: status, error counts, timestamps       |
| `messages(x)`         | One row per error: column, row, rule, message             |
| `validation_status(x)`| Compact status table (status, ok, n_schema_errors, …)    |

```{r inspect_results}
# Summary per table
results(ds)

# Detailed per-error messages (empty for a passing dataset)
msgs <- messages(ds)
head(msgs)
```

### Drilling into Errors with `inspect()`

`messages()` gives you a row per error with an `id` column. When you want the
full story behind a specific error — the exact cell value, the row in context,
the JSON Schema constraint that failed, or the rows that violated a rule —
call `inspect(x, id = N)`.

```
results(x)     → per-table pass/fail summary
messages(x)    → per-error table — note the `id` of the error you care about
inspect(x, id) → deep detail for that error: row context, schema clause, failing rows
```

`inspect()` returns a flat data frame. Columns differ by error type:

| Source   | Key columns added by `inspect()`                                              |
|----------|-------------------------------------------------------------------------------|
| `schema` | `context_*` (the actual row values), `schema_keyword`, `schema_message`       |
| `rule`   | `rule_id`, `failing_row_count`, `failing_*` (preview of violating rows)       |

```{r inspect_example}
# Load a dataset that has schema errors
err_file <- system.file("extdata", "clinical_data_error_schema.csv", package = "DTAtools")
dta_err  <- read_dta_from_yaml(
  system.file("extdata", "clinical_dta.yaml", package = "DTAtools")
)
dta_err  <- load_file(dta_err, "clinical_data", file = err_file)
dta_err  <- check(dta_err, quiet = TRUE)

# Step 1: see which messages exist and note their ids
msgs_err <- messages(datasets(dta_err, "clinical_data"))
head(msgs_err[, c("id", "source", "row", "column", "message")], 8)
```

```{r inspect_schema}
# Step 2: inspect a schema error by id — returns the failing row in context
detail <- inspect(datasets(dta_err, "clinical_data"), id = 1)
# Key fields for schema errors:
detail[, c("id", "message", "type", "why",
           "context_.row", "context_SUBJECT_ID", "context_VISIT",
           "schema_keyword", "schema_message")]
```

```{r inspect_rule}
# Rule errors: inspect shows which rows violated the rule
rule_file <- system.file("extdata", "clinical_data_error_rules.csv", package = "DTAtools")
dta_rule  <- read_dta_from_yaml(
  system.file("extdata", "clinical_dta.yaml", package = "DTAtools")
)
dta_rule  <- load_file(dta_rule, "clinical_data", file = rule_file)
dta_rule  <- check(dta_rule, quiet = TRUE)

msgs_rule <- messages(datasets(dta_rule, "clinical_data"))
msgs_rule[, c("id", "rule_id", "message")]

# Drill into the first rule violation
detail_rule <- inspect(datasets(dta_rule, "clinical_data"), id = 1)
detail_rule[, c("id", "rule_id", "type", "why",
                "failing_row_count", "failing_.row",
                "failing_SUBJECT_ID", "failing_VISIT")]
```

### Example: Schema Errors

The package ships with a CSV that deliberately violates several column
constraints. The error file has:

- A missing `HEIGHT` column (required by spec, since `nullable: false`)
- An invalid `GENDER` value `"Alien"` (not in the allowed list)
- A non-numeric `BMI` cell (column is `SAS Num`)
- An invalid `STUDYID` `"9999-9999"` (not in allowed values)
- A `VISIT` value that exceeds the `length: 12` constraint

```{r schema_errors}
dta_obj_e  <- read_dta_from_yaml(
  system.file("extdata", "clinical_dta.yaml", package = "DTAtools")
)
err_file   <- system.file("extdata", "clinical_data_error_schema.csv", package = "DTAtools")
raw_err    <- read.csv(err_file)

ds_err <- DTADataSetTabular(
  name   = "clinical_data",
  specs  = specs(datasets(dta_obj_e, "clinical_data")),
  tables = list(clinical_data = raw_err)
)
ds_err <- check(ds_err)

# See which tables failed
results(ds_err)

# Get the individual error messages (first 10)
msgs_err <- messages(ds_err)
head(msgs_err[, c("target", "source", "row", "column", "message")], 10)
```

### Example: Schema Rule Violations

This CSV has valid column values but violates semantic rules:

- `CONSENT = "YES"` but `CONSENT_DATE` is empty (`rule_dependency_example`)
- `VISIT = "V03"` but `STATUS = "DROPPED"` (`rule_equal_example`)
- `AGE = 70`, outside the allowed range 18–65 (`rule_range_example`)
- `SUBJECT_ID + VISIT` combination is not unique (`rule_unique_example`)
- `WEIGHT < 5` when `VISIT = V03` (`check_col_condition_example`)

```{r rule_errors}
dta_obj_r  <- read_dta_from_yaml(
  system.file("extdata", "clinical_dta.yaml", package = "DTAtools")
)
rule_file  <- system.file("extdata", "clinical_data_error_rules.csv", package = "DTAtools")
raw_rule   <- read.csv(rule_file)

ds_rule <- DTADataSetTabular(
  name   = "clinical_data",
  specs  = specs(datasets(dta_obj_r, "clinical_data")),
  tables = list(clinical_data = raw_rule)
)
ds_rule <- check(ds_rule)

results(ds_rule)
messages(ds_rule)[, c("target", "source", "rule_id", "message")]
```

---

## Schema Rules

Schema rules are defined alongside column specs and express inter-column
constraints that cannot be captured by JSON Schema alone.

### Rule Types

#### `col_condition` — If/Then Logic

The most versatile rule type. When `condition` is satisfied for a row, the
`then` block is enforced on that same row.

**Supported operators in both `condition` and `then`:**

| Operator        | Applies to        | Example                          |
|-----------------|-------------------|----------------------------------|
| `equals`        | string / number   | `equals: "YES"`                  |
| `not_equals`    | string / number   | `not_equals: "DROPPED"`          |
| `empty`         | any (`true/false`)| `empty: true`                    |
| `in`            | list              | `in: [V01, V02, EOT]`            |
| `not_in`        | list              | `not_in: [REJECTED]`             |
| `pattern`       | string regex      | `pattern: "^SUBJ[0-9]{4}$"`     |
| `greater`       | number            | `greater: 0`                     |
| `greater_equal` | number            | `greater_equal: 18`              |
| `less`          | number            | `less: 200`                      |
| `less_equal`    | number            | `less_equal: 65`                 |
| `min` / `max`   | number range      | `min: 10` + `max: 100`           |

```yaml
rules:
  # IF consent was given, consent date must be filled in
  - id: consent_requires_date
    type: col_condition
    condition:
      CONSENT:
        equals: "YES"
    then:
      CONSENT_DATE:
        empty: false

  # IF visit is V03, status must be COMPLETED
  - id: v03_must_be_completed
    type: col_condition
    condition:
      VISIT:
        equals: V03
    then:
      STATUS:
        equals: COMPLETED

  # IF sample has no reason for test not done, result must be present
  - id: result_when_no_reason
    type: col_condition
    condition:
      GFREASND:
        empty: true
    then:
      GFORRES:
        empty: false

  # Multiple conditions and multiple then-clauses are fully supported
  - id: v03_full_check
    type: col_condition
    condition:
      VISIT:
        equals: V03
    then:
      STATUS:
        equals: COMPLETED
      AGE:
        min: 18
        max: 65
      WEIGHT:
        greater_equal: 5
```

#### `col_range` — Numeric Range Check

Enforces that all values in a numeric column fall within `[min, max]`.

```yaml
rules:
  - id: age_range
    type: col_range
    column: AGE
    min: 18
    max: 65
```

#### `col_unique` — Uniqueness Constraint

Enforces that all combinations of values across one or more columns are unique.
Specifying multiple columns checks the combined key for uniqueness.

```yaml
rules:
  # Single column uniqueness
  - id: subject_unique
    type: col_unique
    column: SUBJECT_ID

  # Composite key uniqueness: SUBJECT_ID + VISIT must be unique per row
  - id: subject_visit_unique
    type: col_unique
    column:
      - SUBJECT_ID
      - VISIT
```

### Defining Rules in R

Rules can also be constructed programmatically:

```{r rules_in_R}
# col_range rule
range_rule <- DTARuleColRange(id = "age_range", columns = "AGE", min = 18, max = 65)

# col_unique rule (composite key)
unique_rule <- DTARuleColUnique(id = "subject_visit_unique", columns = c("SUBJECT_ID", "VISIT"))

# col_condition rule: IF CONSENT == "YES" THEN CONSENT_DATE must not be empty
condition_rule <- DTARuleColCondition(
  id        = "consent_requires_date",
  condition = list(CONSENT = list(equals = "YES")),
  then      = list(CONSENT_DATE = list(empty = FALSE))
)

# Collect rules
rule_collection <- DTARuleCollection(
  rules = list(range_rule, unique_rule, condition_rule)
)
print(rule_collection)
```

```{r specs_with_rules}
# Attach rules to a DTAColumnSpecCollection
studyid  <- DTAColumnSpec(id = "STUDYID",      type = "SAS Char", nullable = FALSE,
                          values = list("STUDY001"))
visit2   <- DTAColumnSpec(id = "VISIT",         type = "SAS Char", nullable = FALSE,
                          values = list("SCREENING", "BASELINE", "WEEK4"))
age2     <- DTAColumnSpec(id = "AGE",           type = "SAS Num",  nullable = FALSE)
consent  <- DTAColumnSpec(id = "CONSENT",       type = "SAS Char", nullable = TRUE,
                          values = list("YES", "NO"))
con_date <- DTAColumnSpec(id = "CONSENT_DATE",  type = "SAS Char", nullable = TRUE)
subjid2  <- DTAColumnSpec(id = "SUBJECT_ID",    type = "SAS Char", nullable = FALSE)

specs_with_rules <- DTAColumnSpecCollection(
  columns = list(
    STUDYID = studyid, VISIT = visit2, AGE = age2,
    CONSENT = consent, CONSENT_DATE = con_date, SUBJECT_ID = subjid2
  ),
  rules = rule_collection
)
print(specs_with_rules)

# Validate data that violates the rules
bad_data <- data.frame(
  STUDYID      = c("STUDY001", "STUDY001"),
  VISIT        = c("SCREENING", "BASELINE"),
  AGE          = c(16, 32),         # 16 violates age_range (min 18)
  CONSENT      = c("YES", "NO"),
  CONSENT_DATE = c(NA, NA),          # CONSENT=YES but no date: violates consent_requires_date
  SUBJECT_ID   = c("P01", "P01")    # duplicate SUBJECT_ID: violates subject_visit_unique when VISIT differs too
)

ds_bad <- DTADataSetTabular(
  name   = "demo_rules",
  specs  = specs_with_rules,
  tables = list(demo = bad_data)
)
ds_bad <- check(ds_bad)

messages(ds_bad)[, c("source", "rule_id", "row", "column", "message")]
```

---

## File-based Workflow

When the DTA defines expected files, use `DTAFileCSV` (or `DTAFileTSV`,
`DTAFileDelim`) as file handlers inside the dataset. The `read_file()` function
reads data through the handler, returning an Arrow Table you can pass directly
to `DTADataSetTabular`.

```{r file_workflow}
# Create a CSV file handler — describes the expected file format
file_handler <- DTAFileCSV(filename = "clinical_data.csv")

# Show what the handler captures
print(file_handler)

# Read the actual file through the handler
csv_path <- system.file("extdata", "clinical_data.csv", package = "DTAtools")
tbl      <- read_file(file_handler, csv_path)
cat("Loaded table class:", class(tbl)[1], "\n")
cat("Dimensions:", nrow(tbl), "x", ncol(tbl), "\n")

# The Arrow Table can be passed directly to DTADataSetTabular
simple_specs <- DTAColumnSpecCollection(columns = list(
  STUDYID = DTAColumnSpec(id = "STUDYID", type = "SAS Char", nullable = FALSE,
                          values = list("1234-5678")),
  VISIT   = DTAColumnSpec(id = "VISIT",   type = "SAS Char", nullable = FALSE,
                          values = list("V01","V02","V03","EOT")),
  AGE     = DTAColumnSpec(id = "AGE",     type = "SAS Num",  nullable = FALSE)
))

# Use the Arrow Table directly in the constructor
ds_from_file <- DTADataSetTabular(
  name   = "clinical_from_file",
  specs  = simple_specs,
  files  = list(file_handler),    # document the expected file
  tables = list(clinical_data = tbl)  # provide the loaded data
)
labels(ds_from_file)
```

For full DTA YAML specs that include file definitions, the file handlers are
embedded in each dataset specification and can be retrieved with `files()`:

```{r file_handlers_from_dta}
dta_spec  <- read_dta_from_yaml(
  system.file("extdata", "clinical_dta.yaml", package = "DTAtools")
)
ds_s      <- datasets(dta_spec, "clinical_data")

# Inspect the file handler defined in the YAML
fh <- files(ds_s, 1)
print(fh)

# Read the file using the handler
csv_path  <- system.file("extdata", "clinical_data.csv", package = "DTAtools")
loaded    <- read_file(fh, csv_path)
cat("Rows loaded:", nrow(loaded), "\n")
```

---

## `DTADataSetFile`: File Presence and Readability Checks

While `DTADataSetTabular` validates the *content* of a delivered file against a
column schema and rules, `DTADataSetFile` answers a more basic question:
**did the expected files actually arrive, and can we open them?**

Use `DTADataSetFile` when a DTS requires that certain non-tabular files (PDFs,
archives, reports, audit logs, raw instrument output) are present and
non-empty — without caring about their internal structure.

`check()` for a `DTADataSetFile` verifies each path:

| Check performed       | Failure message                                  |
|-----------------------|--------------------------------------------------|
| File exists           | `"File '…' not found."`                         |
| File is non-empty     | `"File '…' is empty."`                          |
| File is readable      | `"File '…' is not readable."`                   |

The same `results()`, `messages()`, and `inspect()` API applies, so you can
report file-level problems alongside tabular schema errors in a unified
validation summary.

```{r dataset_file}
# Paths to existing files (pass)
delivered_csv <- system.file("extdata", "clinical_data.csv", package = "DTAtools")
delivered_tsv <- system.file("extdata", "gf_data_small_smirna.tsv", package = "DTAtools")

ds_file <- DTADataSetFile(
  name  = "delivery_check",
  paths = c(delivered_csv, delivered_tsv)
)
ds_file <- check(ds_file)
results(ds_file)     # both files pass
```

```{r dataset_file_missing}
# Introduce a missing file to trigger a failure
ds_missing <- DTADataSetFile(
  name  = "delivery_check_fail",
  paths = c(delivered_csv, "/nonexistent/path/report.pdf")
)
ds_missing <- check(ds_missing)

results(ds_missing)          # one row fails

# The message explains exactly why
messages(ds_missing)[, c("target", "rule_id", "message")]
```

### Mixing `DTADataSetTabular` and `DTADataSetFile` in one `DTA`

A single `DTA` object can contain datasets of both types, letting you validate
the full delivery — column-level content *and* file presence — in one `check()`:

```{r dta_mixed_types, eval=FALSE}
# Tabular dataset (column spec validation)
ds_tab <- DTADataSetTabular(
  name   = "clinical_data",
  specs  = import_specs_from_yaml(
    system.file("extdata", "gf_dataset.yaml", package = "DTAtools")
  ),
  tables = list(clinical_data = read.csv(delivered_csv))
)

# File dataset (presence check)
ds_fi <- DTADataSetFile(
  name  = "delivery_manifest",
  paths = c(delivered_csv, delivered_tsv)
)

dta_mixed <- DTA(
  datasets = list(ds_tab, ds_fi),
  metadata = DTAMetaData(title = "Mixed DTA", version = "1.0")
)

dta_mixed <- check(dta_mixed)
results(dta_mixed)   # one row per dataset
```

---

## The Full DTA Object

### `DTAMetaData`

`DTAMetaData` captures all the administrative information about a DTA:
title, version, contacts (receiver and supplier), and transmission details.

```{r metadata}
# Minimal metadata
meta_simple <- DTAMetaData(title = "My Clinical Data Transmission", version = "1.0")
print(meta_simple)

# Full metadata with contacts and transmission schedule
meta_full <- DTAMetaData(
  title    = "Clinical Genomics Data Transmission",
  version  = "2.0",
  date     = as.Date("2026-01-15"),
  header   = "Boehringer Ingelheim",
  receiver = list(
    affiliation = list(name = "External CRO", country = "Germany"),
    contacts = list(
      list(name = "Alice Smith",  role = "Lead Data Manager",
           email = "alice@cro.com", signature = TRUE, reviewer = TRUE),
      list(name = "Bob Johnson",  role = "Bioinformatician",
           email = "bob@cro.com",   signature = TRUE)
    )
  ),
  supplier = list(
    affiliation = list(name = "Pharma Company", country = "USA"),
    contacts = list(
      list(name = "Emily Turner", role = "Senior Data Manager",
           email = "emily@pharma.com", signature = TRUE)
    )
  ),
  transmission = list(
    type                = "Secure SFTP",
    frequency           = "One-time",
    notification        = "Email",
    date_first_transfer = as.Date("2026-02-01"),
    date_last_transfer  = as.Date("2026-03-31")
  )
)
print(meta_full)
```

### Constructing a `DTA` Object

The `DTA` class holds one or more `DTADataSet` objects together with metadata.

```{r dta_object}
# Build from scratch
ds1 <- create_example_DTADataSetTabular(2)  # "demographics" with data
ds2 <- create_example_DTADataSetTabular(3)  # "vitals" with data

dta_obj <- DTA(
  datasets = list(ds1, ds2),
  metadata = meta_simple
)
print(dta_obj)

# Access datasets by name or index
datasets(dta_obj, "demographics")
dta_obj[["vitals"]]
dta_obj[c(1, 2)]        # returns a list of both datasets
```

### Loading a Full DTA from YAML

The most common production workflow is to define the entire DTA — metadata,
datasets, column specs, rules, and file patterns — in a single YAML file and
load it with `read_dta_from_yaml()`.

```{r read_dta_yaml}
dta_file      <- system.file("extdata", "clinical_dta.yaml", package = "DTAtools")
dta_from_file <- read_dta_from_yaml(dta_file)
print(dta_from_file)

# Inspect metadata
metadata(dta_from_file)
```

### Validating a Full DTA

After loading the DTA definition, load data and validate all datasets:

```{r validate_full_dta}
# Get the specs from the DTA
ds_spec_full <- datasets(dta_from_file, "clinical_data")
clinical_sps <- specs(ds_spec_full)

# Load the CSV data
csv_path_full <- system.file("extdata", "clinical_data.csv", package = "DTAtools")
raw_full      <- read.csv(csv_path_full)

# Create DTADataSetTabular with specs + data
ds_full <- DTADataSetTabular(
  name   = "clinical_data",
  specs  = clinical_sps,
  tables = list(clinical_data = raw_full)
)

# Validate
ds_full <- check(ds_full)
results(ds_full)
```

---

## Exporting Data and Documentation

### Write Validated Tables to File

`write_table_to_file()` writes a named table from a `DTADataSetTabular` to disk.
It optionally computes and writes an MD5 checksum sidecar file.

```{r write_table}
# Validate the clinical dataset
csv_path_w    <- system.file("extdata", "clinical_data.csv", package = "DTAtools")
dta_obj_w     <- read_dta_from_yaml(
  system.file("extdata", "clinical_dta.yaml", package = "DTAtools")
)
ds_w <- DTADataSetTabular(
  name   = "clinical_data",
  specs  = specs(datasets(dta_obj_w, "clinical_data")),
  tables = list(clinical_data = read.csv(csv_path_w))
)
ds_w <- check(ds_w, quiet = TRUE)

# Write to TSV (tab-separated, sorted by SUBJECT_ID)
out_tsv <- tempfile(fileext = ".tsv")
write_table_to_file(
  DTADataSetTabular = ds_w,
  table             = "clinical_data",
  filename          = out_tsv,
  sep               = "\t",
  arrange_by        = "SUBJECT_ID",
  get_md5sum        = FALSE
)
# Confirm output
file.exists(out_tsv)
nrow(read.table(out_tsv, header = TRUE, sep = "\t"))
unlink(out_tsv)
```

Key arguments for `write_table_to_file()`:

| Argument              | Default    | Description                                             |
|-----------------------|------------|---------------------------------------------------------|
| `sep`                 | `"\t"`     | Field separator (`","` for CSV, `"\t"` for TSV)        |
| `arrange_by`          | `"all"`    | A single column name to sort by, `"all"` to sort by all columns, or `NULL` to skip sorting |
| `compression`         | `"none"`   | `"none"` or `"gzip"`                                   |
| `get_md5sum`          | `TRUE`     | Compute and print the MD5 checksum of the written file  |
| `write_md5sum_to_file`| `TRUE`     | Write MD5 + row/column count to a `.md5` sidecar file  |
| `overwrite`           | `TRUE`     | Whether to overwrite an existing file                  |

### Export Spec Table to Word

Generate a formatted Word table of all column specifications — ideal for
pasting directly into a DTA document.

```{r export_word, eval=FALSE}
# gf_specs was loaded earlier with import_specs_from_yaml()
# Export with default columns: Variable Name, Label, Type, Length, Format, Nullable, Description
docx_out <- tempfile(fileext = ".docx")
export_specs_table(gf_specs, file = docx_out, overwrite = TRUE)
```

Two column layouts are supported:

```{r export_word_layouts, eval=FALSE}
# Layout 1: without Length column
export_specs_table(
  gf_specs,
  file      = "spec_table_short.docx",
  overwrite = TRUE,
  colnames  = c("Variable Name", "Variable Label", "Type", "Format", "Nullable", "Description")
)

# Layout 2: with Length column (default)
export_specs_table(
  gf_specs,
  file      = "spec_table_full.docx",
  overwrite = TRUE,
  colnames  = c("Variable Name", "Variable Label", "Type", "Length", "Format", "Nullable", "Description")
)
```

### Export Column Values Table to Word

Generate a Word table of all allowed values for a specific column — useful for
listing controlled vocabulary in a DTA document.

```{r export_values, eval=FALSE}
# Export all allowed VISIT values for the clinical dataset
docx_values     <- tempfile(fileext = ".docx")
dta_obj_export  <- read_dta_from_yaml(
  system.file("extdata", "clinical_dta.yaml", package = "DTAtools")
)
clinical_specs_export <- specs(datasets(dta_obj_export, "clinical_data"))

export_column_value_table(
  DTAColumnSpecCollection = clinical_specs_export,
  file = docx_values,
  id   = "VISIT"
)
```

---

## YAML Format Reference

### Column Specification

```yaml
columns:
  # Minimal example — id + type + nullable
  - id: STUDYID
    label: Study Identifier
    type: SAS Char
    nullable: false
    length: 10
    values:
      - 1234-5678
    description: Unique identifier for the study

  # Closed value list
  - id: GENDER
    label: Gender
    type: SAS Char
    nullable: false
    values:
      - Male
      - Female
      - Other

  # Regex pattern (mutually exclusive with values)
  - id: SUBJECT_ID
    label: Subject Identifier
    type: SAS Char
    nullable: false
    pattern: "^SUBJ[0-9]{4}$"

  # Numeric column, nullable
  - id: AGE
    label: Age (years)
    type: SAS Num
    nullable: false
    description: Age in whole years at baseline

  # Nullable character (empty/NA is allowed)
  - id: AE_TERM
    label: Adverse Event Term
    type: SAS Char
    nullable: true
    description: Adverse event verbatim term, empty if no event
```

**Supported types:** `SAS Char`, `SAS Num`, `SAS Int`

**Nullable values:** `true`/`false` or `Yes`/`No`

### Schema Rules

```yaml
rules:
  # ── col_condition ──────────────────────────────────────────────────────────
  # IF CONSENT == "YES" THEN CONSENT_DATE must not be empty
  - id: consent_requires_date
    type: col_condition
    condition:
      CONSENT:
        equals: "YES"
    then:
      CONSENT_DATE:
        empty: false

  # IF VISIT == V03 THEN STATUS must be COMPLETED
  - id: v03_completed
    type: col_condition
    condition:
      VISIT:
        equals: V03
    then:
      STATUS:
        equals: COMPLETED

  # IF GFREASND is populated THEN GFORRES must be empty (mutual exclusivity)
  - id: mutual_exclusivity
    type: col_condition
    condition:
      GFREASND:
        empty: false
    then:
      GFORRES:
        empty: true

  # Multiple then-clauses: age range, weight lower bound, status check
  - id: v03_full_checks
    type: col_condition
    condition:
      VISIT:
        equals: V03
    then:
      STATUS:
        equals: COMPLETED
      AGE:
        min: 18
        max: 65
      WEIGHT:
        greater_equal: 5

  # Pattern matching in condition
  - id: internal_supplier
    type: col_condition
    condition:
      GFREFID:
        pattern: "^Internal_[0-9]+$"
    then:
      GFNAM:
        equals: "Internal"

  # ── col_range ──────────────────────────────────────────────────────────────
  # AGE must be between 18 and 65 (inclusive)
  - id: age_range
    type: col_range
    column: AGE
    min: 18
    max: 65

  # ── col_unique ─────────────────────────────────────────────────────────────
  # SUBJECT_ID must be globally unique
  - id: subject_unique
    type: col_unique
    column: SUBJECT_ID

  # The combination of SUBJECT_ID + VISIT must be unique
  - id: subject_visit_unique
    type: col_unique
    column:
      - SUBJECT_ID
      - VISIT
```

### Full DTA YAML Structure

A complete DTA YAML file ties together metadata, dataset definitions (with file
patterns), column specs, and rules in a single document:

```yaml
metadata:
  title: Clinical Data Transmission Agreement
  version: "1.0"
  date: 2026-01-15
  header: Boehringer Ingelheim

  receiver:
    affiliation:
      name: External CRO
      country: Germany
    contacts:
      - name: Alice Smith
        role: Lead Data Manager
        email: alice@cro.com
        signature: true
        reviewer: true

  supplier:
    affiliation:
      name: Pharma Company
      country: USA
    contacts:
      - name: Emily Turner
        role: Senior Data Manager
        email: emily@pharma.com
        signature: true

  transmission:
    type: Secure SFTP
    frequency: one-time
    notification: email
    date_first_transfer: 2026-02-01
    date_last_transfer: 2026-03-31

datasets:
  - name: clinical_data
    type: tabular
    description: Clinical subject-level data

    files:
      filename: "clinical_data.*.csv$"
      pattern: Yes
      number_of_files: 1
      type: csv

    columns:
      - id: STUDYID
        label: Study Identifier
        type: SAS Char
        nullable: false
        length: 10
        values:
          - 1234-5678
      - id: VISIT
        label: Visit Code
        type: SAS Char
        length: 12
        nullable: false
        values:
          - V01
          - V02
          - V03
          - EOT
      - id: AGE
        label: Age (years)
        type: SAS Num
        nullable: false
      - id: SUBJECT_ID
        label: Subject Identifier
        type: SAS Char
        nullable: false
        pattern: "^SUBJ[0-9]{4}$"

    rules:
      - id: age_range
        type: col_range
        column: AGE
        min: 18
        max: 65
      - id: subject_visit_unique
        type: col_unique
        column:
          - SUBJECT_ID
          - VISIT
      - id: v03_completed
        type: col_condition
        condition:
          VISIT:
            equals: V03
        then:
          STATUS:
            equals: COMPLETED
```

---

## API Quick Reference

### Core Classes

| Class                     | Purpose                                                     |
|---------------------------|-------------------------------------------------------------|
| `DTA`                     | Top-level container: metadata + multiple datasets           |
| `DTAMetaData`             | Agreement metadata (title, version, contacts, transmission) |
| `DTADataSet`              | Abstract base class for all dataset types                   |
| `DTADataSetTabular`       | Dataset with column specs, rules, and data tables           |
| `DTADataSetFile`          | Dataset that checks file presence and readability only      |
| `DTAColumnSpecCollection` | Named list of column specs + optional rule collection       |
| `DTAColumnSpec`           | Single column definition (type, nullable, values, pattern)  |
| `DTARuleCollection`       | Container of DTARule objects                                |
| `DTARuleColCondition`     | If/then cross-column rule                                   |
| `DTARuleColRange`         | Numeric range constraint for a column                       |
| `DTARuleColUnique`        | Uniqueness constraint (single or composite key)             |
| `DTAFileCSV`              | CSV file handler for `read_file()`                          |
| `DTAFileTSV`              | TSV file handler for `read_file()`                          |

### Key Functions

| Function                    | Description                                                  |
|-----------------------------|--------------------------------------------------------------|
| `read_dta_from_yaml(file)`  | Load a full DTA from YAML                                    |
| `import_specs_from_yaml(file)` | Load column specs + rules from a standalone YAML         |
| `columns_specs_from_word(file)` | Import column specs from a Word table                   |
| `load_file(dta, dataset, file)` | Read a data file into a dataset using its YAML-defined handler |
| `read_file(handler, file)`  | Read a file using a file handler (`DTAFileCSV`, etc.)        |
| `check(x)`                  | Validate all datasets / tables; returns updated object       |
| `results(x)`                | Summary table: status, error counts per table                |
| `messages(x)`               | Detailed error table: row, column, rule, message, `id`       |
| `inspect(x, id)`            | Deep detail for error `id`: row context, schema clause, failing rows |
| `validation_status(x)`      | Compact status data frame                                    |
| `validation_errors(x, table)` | Full raw error output for one table                       |
| `datasets(x, name)`         | Extract dataset(s) from a DTA object                        |
| `tables(x, i)`              | Extract table(s) from a DTADataSetTabular                    |
| `get_table(x, id)`          | Extract a single Arrow Table by name or index               |
| `labels(x)`                 | Names of all tables in a DTADataSetTabular                   |
| `specs(x)`                  | DTAColumnSpecCollection from a DTADataSetTabular             |
| `colspec(x, id)`            | Single DTAColumnSpec by column ID                           |
| `rules(x)`                  | DTARuleCollection from a collection or dataset              |
| `metadata(x)`               | DTAMetaData from a DTA object                               |
| `write_table_to_file(…)`    | Write a validated table to disk (TSV/CSV, gzip, MD5)        |
| `write_columns_to_yaml(x, file)` | Serialise specs to YAML                               |
| `write_columns_to_json(x, file)` | Serialise specs to JSON                               |
| `export_specs_table(x, file)` | Export spec table to Word                               |
| `export_column_value_table(x, file, id)` | Export value list for a column to Word         |
| `as_json_schema(x)`         | Convert specs to JSON Schema string                         |

---

## Credits

`DTAtools` was developed at Boehringer Ingelheim by:

- **Thomas Schwarzl** (aut, cre)
- **Daniel Schreyer** (aut, ctb)

The package is open-sourced to help external and internal collaborators validate
data before file transfer, following the DevOps principle of detecting errors
fast and automatically.
